---
title: "AusTraits"
format: html
editor_options: 
  chunk_output_type: console
---

# Introduction & Goals

(daniel)
- Introduce AusTraits as an example of a big data
- Refresh and practice skills in data manipulation and sumamrising
- learn some fancy tricks: nests, joins, 

https://doi.org/10.1038/s41597-021-01006-6


# Setup 

## Accessing AusTraits

https://zenodo.org/record/7583087/files/austraits-4.1.0.rds?download=1

save to downloads folder in this project

## Loading data

Use function `readRDS` to read in the data

```{r}
austraits <- readRDS("downloads/austraits-4.1.0.rds")
```

## Packages
(daniel)
To analyse, goingto use the tidyverse package

- links about tidyvesre

# Exploring the data
(daniel)
AusTraits object is a list

## Dictionary

Data structure described in dictionary

What is???

## Lists and  Tables (tibbles)

```{r}
names(austraits)
```

Access elements withins

```{r}
austraits[["traits"]]
```

Most parts are tables (tibbles)

Challenge: 

How many rows in each of these tables XXXX:

```{r}

```


## Working with tables

Chance to practice working with and summarising data

dplyr functions

- filter
- select
- mutate
- summarise
- group_by

Challenges

1. How many taxa and traits are included?

```{r}

```

Answer: 

2. From taxa table, how many accepted species?

```{r}

```

Answer: 


2. How many trait records per taxa? Arrange taxa from most to least. What are top 5 taxa?

```{r}

```

Answer: 

3. 

xx




## Make plots with ggplot

Basic ggplot formulas

```{r}

```

Challenges

Plot XXX
```{r}

```

Recreate Figure 2a.

```{r}

```



## Nest



## Pivot wider


## Joining tables


